\section{The Pratyabhij\~n\=a Creative Engine}\label{sec:engine}

The engine is a deterministic pipeline of seven operators plus two sleep-style consolidation cycles, all implemented in \texttt{src/pce/}. We summarise each operator's signature here and refer to Appendix~\ref{app:operator_specs} for the full equations and Appendix~\ref{app:consolidation} for the consolidation cycles.

\subsection{Substrate}
Two substrate primitives ground the cascade in real computation: a sentence-embedder $\phi_{\text{embed}}$ (\texttt{sentence-transformers/all-MiniLM-L6-v2}, $d=384$) and a generative LM. v0.3 supports two LM substrates behind a unified \texttt{GeneratorProtocol} interface: a local causal LM $\pi_{\text{LM}}$ (\texttt{Qwen/Qwen2-1.5B-Instruct}, autodetected device CUDA $\rightarrow$ MPS $\rightarrow$ CPU, \texttt{float16} where supported), and Anthropic's Claude Haiku via a \emph{clean CLI substrate}: every Haiku call is dispatched through \texttt{claude --print --tools "" --strict-mcp-config --disable-slash-commands --setting-sources='' --no-session-persistence} as a subprocess, run inside a scrubbed \texttt{HOME} that contains only the keychain symlink (macOS) or \texttt{.config/claude/} symlink (Linux) needed for OAuth. An explicit env allow-list (\texttt{ENV\_ALLOWLIST}) replaces any blind \texttt{os.environ} inheritance, and a per-process \texttt{IntegrityProbe} (\texttt{pce.substrate.integrity}) periodically asks the inner subprocess what plugins/skills/MCP tools it has access to and \texttt{LEAKAGE\_REGEX}-scans the response (negation-aware, so ``I have no plugins'' does not trigger). Sampling is deterministic given a seed under both substrates.

\subsection{Operators}

\begin{itemize}[leftmargin=1.2em]
\item $\textsc{Cit}(\text{prompt}, \tau, \text{seed}) \to (\text{tokens}, \log p, \text{embedding})$. Direct call to $\pi_{\text{LM}}.\text{generate}$. v0.3 plumbs the \emph{cit-temperature} $\tau_{\text{cit}}$ through to the \iast{icch\=a} fan-out posterior (see below) so the substrate's exploration--exploitation knob actually reaches \iast{j\~n\=ana}'s evidence accumulator.
\item $\textsc{\=Ananda}(\text{candidate}, \text{constraint}, \text{retrieval-set}) \to r \in [0,1]$. A weighted aggregate of (i) constraint-cosine, (ii) retrieval-set diversity, (iii) form-fidelity heuristics (line / syllable proxies for poetry).
\item $\textsc{Icch\=a}(\text{prompt}, K, \text{constraint}, \tau_{\text{cit}}) \to \{c_1, \dots, c_K\}$. $K$ candidate generations under a constraint-injected prompt; the parity sampler's $\tau_{\text{base}}$ is now scaled by $\tau_{\text{cit}}$ and recorded on each candidate's \texttt{sampler} dict so the audit log captures the temperature used.
\item $\textsc{Apohana}(\{c_i\}, \text{must-avoid}, \text{Hopfield}) \to \{a_i\}$. Contrastive embedding score with v0.3's optional Hopfield warm-start: $a_i = 1 - \max_j \cos(\phi(c_i), \phi(\text{must-avoid}_j)) + w \cdot \text{HopfieldQuery}(\phi(c_i))$. The Hopfield bonus reads a per-domain \iast{\=alayavij\~n\=ana} (\texttt{HopfieldStore} from \texttt{pce.active\_inference.hopfield}) under SWS/REM consolidation modes.
\item $\textsc{J\~n\=ana}(\{c_i, a_i\}, \text{aspects}, \pi_{\text{aspect}}) \to (i^*, \Delta F, p)$. v0.3 adds an \emph{aspect-conditioned} reduction target: when aspects are supplied, BMR reduces over the $|aspects|$ candidate models that boost in-support candidates by $1 + \text{boost}$, leaving out-of-support candidates at $1$. This produces a non-degenerate $\Delta F$ that responds to aspect priors $\pi_{\text{aspect}}$ derived from the Hopfield warm-start. The selected $i^*$ is the posterior maximum; $\Delta F$ is recorded and feeds the event-gated commit policy.
\item $\textsc{Kriy\=a}(c_{i^*}, \text{mode}) \to \text{surface}$. Three modes: \texttt{verbatim} (return the candidate unchanged), \texttt{polish} (deterministic detokenisation), \texttt{claude\_polish} (Claude Code's editing surface, opt-in).
\item $\textsc{Vimar\'sa}(\text{surface}, \text{aspects}, \text{retrieval-set}, \Delta F) \to (\text{event}, \text{novelty}, \text{brief})$. v0.3 makes \iast{vimar\'sa} a \emph{causal} step on the cascade: an event fires when $|\Delta F| \geq \texttt{DEFAULT\_DELTA\_F\_THRESHOLD}$ (default 0.01) \emph{and} aspect-multiplicity is detected. When the event fires the cascade commits the revision; otherwise it commits the draft. A \emph{shadow} revision is always generated and scored regardless of commit outcome, so H8.v3 (revision-vs-draft within cascade) is measurable on every item. The committed surface is then \texttt{consolidate}d into the Hopfield store under either REM (append) or SWS (k-means consolidate) mode.
\end{itemize}

\subsection{Cascade}
The cascade in \texttt{src/pce/cascade.py} composes the seven operators in order. The state object \texttt{CascadeState} records \emph{both} \texttt{surface\_draft} and \texttt{surface\_revision}, a \texttt{committed} field (\texttt{"draft"} or \texttt{"revision"}), the \texttt{commit\_policy} that produced it (one of \texttt{"event\_gated"}, \texttt{"always\_revise"}, \texttt{"always\_draft"}), the per-item \texttt{FreeEnergyBudget} ledger, and every $\Delta F$ along the way. The cascade is deterministic: identical inputs and seed produce byte-identical outputs.

\subsection{Consolidation}
Two cycles operate over the Hopfield store: \textsc{ConsolidateSWS} (k-means with k chosen by elbow-statistic) for slow-wave-sleep abstraction; \textsc{ConsolidateREM} (Metropolis re-sampling that perturbs and re-stores) for rare-attractor maintenance. Both cycles are exposed as MCP tools so a downstream agent can request consolidation explicitly.
