\section{Methods}\label{sec:methods}

\subsection{Substrate: from OAuth-bound API (v0.3) to Bedrock pilot (v0.4)}

The v0.3 four-arm matrix ran against the OAuth-bound Anthropic API through the \texttt{claude --print} CLI in a bytewise-isolated subprocess (subprocess flag isolation, scrubbed \texttt{HOME} that selectively symlinks the macOS keychain or \texttt{\textasciitilde/.config/claude/} for Linux, per-item \texttt{IntegrityProbe} on \texttt{(env\_hash, flags\_hash)} that halts on leakage) --- the design discharged the v0.2 Claude-Code-context contamination critique \citep{PratyakshaContextEng2025}. v0.4 keeps that substrate isolation byte-for-byte but switches the \emph{routing target} from the OAuth-bound API to AWS Bedrock, with the model identifiers \texttt{global.anthropic.claude-haiku-4-5-20251001-v1:0} for the cascade scorer and \texttt{global.anthropic.claude-sonnet-4-5-20250929-v1:0} for the judge. The substrate change was driven by quota-availability constraints during the Phase 7 pilot (the OAuth substrate had per-account session caps that would not have admitted a 1,277-call run); ADR-006 (\texttt{docs/adr/v0.4/ADR-006-haiku-rate-limit-error.md}) documents the routing target switch and the rate-limit handling. ADR-007 (\texttt{docs/adr/v0.4/ADR-007-sdk-removal.md}) documents the parallel removal of the (deprecated) Anthropic Python SDK code path: the cascade now exclusively uses the CLI substrate, which gives the architecture a single auth chain across Cursor, Claude Code, the standalone CLI, and the Bedrock pilot (\S\ref{sec:substrate}).

\subsection{Arms (the v0.3 four-arm matrix, retained)}

The four-arm matrix from v0.3 carries forward unchanged in its arm definitions, since the v0.2 confounds it was designed to control are still load-bearing. The first arm, \texttt{haiku\_bare} (no PCE), invokes Claude Haiku through the v0.4 isolated CLI substrate (\S\ref{sec:engine}) as a single sampler call with no cascade, and serves as the architecture-versus-nothing primary control. The second arm, \texttt{haiku\_cascade} (with PCE), routes the same Haiku endpoint through the v0.4 cascade with \texttt{commit\_policy="event\_gated"}, $K{=}3$, \texttt{max\_tokens}${}=200$, the active-inference uplift (aspect-conditioned BMR, Hopfield warm-start, plumbed \iast{cit}-temperature, per-item \texttt{FreeEnergyBudget}), and an always-shadow revision that lets H8a be measurable on every item. The third arm, \texttt{haiku\_bare\_2K\_scorer} (no PCE, $+K$ compute), is a single pass with \iast{icch\=a} fanning out $K'=2K$ candidates and the same \iast{j\~n\=ana} BMR scorer the cascade uses (\texttt{commit\_policy="always\_draft"}); it spends roughly the same Haiku-token budget as the cascade but denies it the second pass and the \iast{vimar\'sa} brief, serving as the H6.v4 extra-compute control. The fourth arm, \texttt{haiku\_generic\_revise\_2pass} (two passes, no \iast{vimar\'sa} brief), runs two passes with \texttt{commit\_policy="always\_revise"} but replaces the \iast{vimar\'sa}-generated brief with a fixed generic creative-revise prompt (\texttt{GENERIC\_REVISE\_BRIEF}), serving as the H7.v4 revision-protocol control.

\subsection{Commit-policy multiplexer (the v0.4 addition)}

A v0.4 redesign of the cascade lifts the commit decision from a single configurable parameter to a multiplexer: the cascade always runs both the draft and the shadow revision, scores both, and then commits the surface chosen by the active commit policy. Four policies are evaluated in parallel arms, each named \texttt{haiku\_cascade\_<policy>}. The \texttt{always\_draft} policy commits the draft and ignores the revision (the degenerate baseline, equivalent to disabling the second pass). The \texttt{always\_revise} policy commits the revision unconditionally. The \texttt{event\_gated} policy commits the revision if and only if the \iast{vimar\'sa} event-disjunction fires (the v0.3 policy carried forward). The \texttt{learned\_gate} policy commits the revision if and only if a small logistic head trained on the H8a paired-revision data (ADR-002) predicts $P(\text{revision better})>0.5$ from a six-feature signature of the draft: composite score, $\Delta F$, aspect-cosine, Hopfield retrieval similarity, the $|\text{revision}-\text{draft}|$ embedding distance, and length.

The multiplexer construction means \emph{all four policies see the same draft and shadow revision} on each item; the only varying choice is which surface gets committed. This makes the H8c policy comparison a fair within-cascade contrast rather than a between-pipeline contrast. Because the same scorer ranks both surfaces, H8c also stress-tests the scorer-vs-judge agreement (H9): a policy that commits the surface the scorer prefers may not commit the surface the \emph{judge} prefers, and the H9 result tells us how often that disagreement matters.

\subsection{Pre-registered hypotheses (v0.4)}\label{sec:hypotheses}

The v0.4 hypothesis stack inherits H1--H4 directly from v0.3 (one per domain, primary cascade-vs-bare contrast), redefines H5 as a fixed-effects pool (ADR-005), retains H6 and H7 as the v0.3 fairness controls, and replaces the single v0.3 H8 with a three-part decomposition (H8a paired-revision quality, H8b gate calibration, H8c commit-policy comparison), plus a new H9 (judge-vs-scorer agreement). All hypotheses are pre-registered in \texttt{docs/SPEC\_v0.4.md} before the Phase 7 Bedrock pilot is run. H1.v4 through H4.v4 each test that \texttt{haiku\_cascade} achieves a higher composite than \texttt{haiku\_bare} on the four respective domains (AUT, poetry-interp, poetry-gen, sci-creativity). H5.v4 (re-registered under ADR-005) pools the four primary contrasts via fixed-effects inverse-variance meta-analysis on per-domain Hedges' $g$, reported as $g$, BCa~95\% CI, and per-domain weights. H6.v4 (extra-compute fairness) tests that \texttt{haiku\_cascade} achieves a higher per-domain composite than \texttt{haiku\_bare\_2K\_scorer}; H7.v4 (revision-protocol fairness) tests the same against \texttt{haiku\_generic\_revise\_2pass}. H8a.v4 (revision quality, paired) tests that within \texttt{haiku\_cascade\_always\_revise}, the paired \texttt{score(surface\_revision)} $-$ \texttt{score(surface\_draft)} is positive over all items. H8b.v4 (gate calibration) evaluates both the event-gated and learned-gate commit policies as binary classifiers (``predict revision is better than draft''), reporting precision, recall, F1, and accuracy on the H8a paired ground truth. H8c.v4 (commit-policy leaderboard) pairwise-compares the four commit policies on paired delta versus \texttt{haiku\_bare}, with paired permutation tests across all six policy pairs. Finally, H9.v4 (judge-vs-scorer agreement) tests that the Sonnet 4.5 judge's per-item delta (cascade $-$ bare) and the Haiku composite scorer's per-item delta correlate at Spearman $\rho > 0$ (one-sided), with a sign-agreement diagnostic on the same paired items.

\subsection{Statistical protocol}

For each per-domain hypothesis we report: paired mean $\Delta$ (raw and length-controlled, where the per-arm linear word-count effect is regressed out before pairing), Hedges' $g$ with small-sample correction (raw and length-controlled), paired permutation $p$ (one-sided directional, exact for $n \le 18$, Monte-Carlo with 50,000 permutations otherwise), Wilcoxon signed-rank $p$ (one-sided), 95\% BCa bootstrap CI of the paired mean (10,000 resamples), Holm--Bonferroni adjusted $p$ across $\{H_1, H_2, H_3, H_4\}$, and a-priori (assumed $g{=}0.5$) and retrospective (observed $g$) statistical power. A hypothesis is \emph{supported} iff its Holm-adjusted $p<0.05$ \emph{and} its BCa CI is strictly positive. H5.v4 is reported as a fixed-effects pool with inverse-variance weights and a 95\% Wald CI on the pooled $g$. H8a is reported as a within-arm paired test with BCa CI. H8b is reported as binary-classifier metrics (precision, recall, F1, accuracy, support). H8c is reported as a leaderboard with paired BCa CIs and a pairwise paired-permutation matrix. H9 is reported as Spearman $\rho$, $p$, and a sign-agreement rate. Every JSON leaf passes through \texttt{\_clean\_json}: non-finite floats become \texttt{null} and the writer uses \texttt{allow\_nan=False}, so \texttt{benchmarks/results\_v0.4/stats.json} is RFC-compliant strict JSON.

\subsection{Reproducibility}

All seeds are recorded; bootstrap and permutation RNGs are seeded by \texttt{--seed} (default 4242). The benchmark driver (\texttt{benchmarks/driver.py}) writes a checkpoint after every call so a re-run resumes exactly from the failure point and runs the per-item \texttt{IntegrityProbe} (cached on \texttt{(env\_hash, flags\_hash)}; halts on leakage unless \texttt{--allow-leakage} is set). The v0.4 audit log lives under \texttt{audit/v0.4/} (cost ledgers per domain plus \texttt{audit/v0.4/cost\_ledger\_merged.json}, \texttt{audit/v0.4/integrity\_probes\_merged.jsonl}, \texttt{audit/v0.4/lit\_verification.jsonl}, \texttt{audit/v0.4/lit\_new\_entries.jsonl}, and \texttt{audit/v0.4/phase8\_preflight.txt}). Every numerical claim in this paper is generated from \texttt{benchmarks/results\_v0.4/stats.json} via \texttt{benchmarks/autoreport.py --version v0.4} and is therefore traceable end-to-end.

\subsection{TRIZ-resolved contradictions (v0.3 carried + v0.4 deltas)}

The five v0.3 TRIZ resolutions (clean substrate vs.\ OAuth, event-gated commit vs.\ measurability, BMR aspect-conditioning, Hopfield in cascade vs.\ purity, free-energy budget vs.\ open-ended cascade) are inherited unchanged. v0.4 introduces three additional architecture-decision records. ADR-001 (best-of-$K$ \iast{cit}-temperature) plumbs the \iast{cit}-temperature through the \iast{icch\=a} fan-out per-candidate, so the $K$-candidate posterior is sampled at a deliberate variance rather than at the single substrate-default temperature. ADR-002 (learned gate) introduces a commit policy that replaces the disjunctive event-gate with a logistic predictor trained on the H8a paired-revision ground truth, tuning the gating threshold against the cascade's own measured revision quality rather than against a hand-coded $\Delta F$ heuristic. ADR-004 (Sonnet judge protocol) issues a separate judge call against \texttt{claude-sonnet-4-5-20250929-v1:0} on every cascade item, scoring the cascade-versus-bare pair under the same axis decomposition the Haiku scorer uses; the H9 agreement test is the principal output. Each ADR is end-to-end traceable from TRIZ card to ADR to operator code to unit test to pilot result.
