\section{Showcase Examples}\label{sec:showcase}

\sloppy

The release ships nine showcase outputs, three per stratum (Sanskrit chandas, English poetry, scientific creativity), each with a full draft, shadow-revision, and commit trace and a per-domain validator output. The Sanskrit items are live cascade traces (markdown-prose answers, not stripped verse surfaces); the English and science items are clean cascade-output surfaces. The full set lives at \nolinkurl{benchmarks/showcase_v0.4/} (and is rendered on the GitHub Pages site). All nine items are produced by the cascade: six are Phase~7 pilot cascade traces; the three Sanskrit chandas items are live cascade outputs regenerated against the prompts in \path{scripts/showcase_specs.toml} with \texttt{source}${}={}$\texttt{live\_cascade}. Each Sanskrit trace ships with the \path{tools/sanskrit_chandas.py} validator's report attached; the current cascade is not chandas-aware, so the validator's pass/fail signal is informational only. Future work includes a chandas-aware composite scorer and a learned chandas-conditional gate; we discuss this scope-cut in \S\ref{sec:discussion} and \S\ref{sec:limitations}. This section walks each item briefly; the full traces are in the artefact directory. Table~\ref{tab:showcase_registry} indexes every showcase slug with its trace provenance, committed surface, and validator status.

\input{sections/_tables/tab_showcase_registry}

\subsection{Sanskrit chandas (3 items, live cascade traces)}

\paragraph{anu\d{s}\d{t}ubh.} A 4$\times$8 \'sloka on the theme of \iast{vimar\'sa} (reflective recognition), generated live by the cascade against the prompt in \path{scripts/showcase_specs.toml} with \texttt{style="sanskrit\_chandas"}. The committed surface is what the cascade emitted; the chandas validator's syllable-count and pattern-match report is included in the trace. We treat the validator as a reporting-only signal: the cascade is not yet conditioned on chandas constraints, so a non-conformant validator output is an honest record of the substrate's current ability rather than a release blocker.

\paragraph{g\=ayatr\={\i}.} A 3$\times$8 invocation to \iast{citi-\'sakti} as the witnessing ground (s\=ak\d{s}in), again live-regenerated. The trace records draft, shadow revision, vimar\'sa event flag, BMR \(\Delta F\), and the validator's pad-by-pad syllable count and any deviation from the canonical pattern.

\paragraph{indravajr\=a.} A 4$\times$11 invocation generated against the indravajr\=a metre prompt. The metre's strict GG\,L\,GG\,L\,L\,G\,L\,G\,G pattern is a hard target the current cascade is not optimised for; the trace ships the cascade's output verbatim, with the validator's reported deviations attached. Whether and how a chandas-aware scorer recovers metre conformance is a future-work falsification target.

\subsection{English poetry (3 items, real Phase 7 cascade traces)}

\paragraph{Imagist haiku (\texttt{p06}, gate worked).} A 5/7/5 imagist haiku (``White feather, iron rails / Sings as the train rushes through / Empty platform sleeps'') was the highest-composite Phase 7 output for \texttt{poetry\_gen} ($\sim\!0.635$). Honest reading: the \iast{vimar\'sa} event \emph{did not} fire on this item, so the cascade committed the draft. The shadow-revision pass produced ``Sings when trains pass through'' --- a slightly different image but the proxy scorer rates it $0.59$ vs.\ the draft's $0.64$, so the event-gate's decision to keep the draft was \emph{correct} by the scorer. This is the cascade's gate working as designed on an item where the draft was already good.

\paragraph{Dickinson slant (\texttt{p07}, gate miss).} A free-verse poem in the Dickinson tradition (slant rhyme, en-dash mid-thought, anaphora). The \iast{vimar\'sa} event did not fire, but the shadow revision (composite $0.567$) clearly beat the draft (composite $0.504$). This is exactly the calibration gap H8b documents: the event-gated commit policy under-fires on this domain. The H8c leaderboard's verdict that \texttt{always\_revise} out-performs \texttt{event\_gated} is driven by items like this.

\paragraph{Pastoral, traditional metre (\texttt{p01}, gate worked).} A traditional pastoral 5/7/5 with a rain-on-tin auditory image. \iast{Vimar\'sa} did not fire. The shadow revision tried to escalate ``sings'' $\to$ ``shrieks through'' and ``deepens to'' $\to$ ``devours the song'' --- a louder register. The proxy scorer rates the draft (0.566) above the revision (0.523), so the event-gate's decision to keep the draft was correct. The revision is shown side-by-side in the showcase for comparison; this is the cascade \emph{not} mistaking volume for craft.

\subsection{Scientific creativity (3 items, real Phase 7 cascade traces)}

\paragraph{Why mathematics works in physics (\texttt{s06}, vimar\'sa fired).} The gold-star case in the current release. Vimar\'sa fired (\texttt{vimarsa\_event=True}), the gate committed the revision, and the revision (composite 0.618) beat the draft (composite 0.587). The revision pass also converted the title from ``A Contrarian View'' to ``The Persistent Mystery'', reframing the essay from polemic to honest puzzle-statement.

\paragraph{Why ice floats --- beyond the textbook (\texttt{s02}, gate miss).} Vimar\'sa did not fire, so the cascade committed the draft (composite 0.535), but the shadow revision scored higher (0.593). The revision changed the title from ``Beyond the Textbook'' to ``Geometry and Thermodynamic Precarity'', a sharper framing that matches the brief. Items like this are the basis for H8c showing \texttt{always\_revise} outperforming \texttt{event\_gated}.

\paragraph{Why galaxies have spiral arms (\texttt{s05}, largest revision uplift).} The largest single-item shadow-revision uplift in the pilot (draft 0.486 $\to$ revision 0.595, $\Delta=+0.110$). Vimar\'sa did not fire so the cascade committed the draft; the shadow revision sharpened ``Framing 1 - The Traffic Jam'' to ``Framing 1 - Density Waves'', upgrading the analogy to its proper physics name (Lin--Shu density-wave theory). The revision is shown for comparison; the gold-star reading is that the cascade \emph{produced} the better surface but failed to commit it. Future work on the learned gate is partly about catching items like this one.
