\section{Limitations and Threats to Validity}\label{sec:limitations}

\paragraph{Pilot is under-powered by design.} The v0.3 pilot was scoped to $n{=}5$/domain ($N_{\text{paired}}{=}20$), $K{=}3$, max\_tokens$=200$ to fit a \$20 Haiku budget. With $n{=}5$ the exact sign-flip permutation test has a hard floor of $p{=}0.0312$ and Holm--Bonferroni with $m{=}4$ floors the smallest possible adjusted $p$ at $0.125$. \emph{No primary hypothesis (H1--H4) can cross the pre-registered Holm $p<0.05$ bar at this sample size}, regardless of effect direction or magnitude. A properly powered run at $n{\sim}20$/domain is required to claim or rule out a stable architectural effect.

\paragraph{The v0.3 pilot reported a directional-negative outcome that v0.4 corrects.} On all four primary domains in v0.3, the \texttt{haiku\_cascade} arm trailed the clean-substrate \texttt{haiku\_bare} arm (Hedges' $g \in [-0.90, -0.22]$); the H5.v3 random-effects pool came out at $g=-0.46$ (95\% CI $[-0.93, +0.01]$), and H6.v3 and H7.v3 also trailed across all four domains. We interpret this finding three ways. First, the v0.2 ``cascade beats bare'' lift, on a clean Haiku CLI substrate that no longer leaks Claude-Code context into the bare control, did not reproduce; this is consistent with the v0.2 adversarial review's specific prediction. Second, against budget-matched (H6.v3) and protocol-matched (H7.v3) controls, the cascade was not preferred on that small sample, and both ``$+K$ compute on a single pass'' and ``a generic 2-pass revise'' --- simpler interventions --- performed at least as well at $n=5$. Third, the within-cascade H8.v3 test was directionally positive ($\Delta=+0.004$, $g=+0.21$) on the $n=3$ items where event-gated commit chose to revise, but $n=3$ was far too small to adjudicate causality. The v0.4 design is the answer to that mismatch: by lifting the commit decision into the four-policy multiplexer, by training a learned gate on the H8a paired-revision data, and by enlarging the within-cascade contrast (H8a) to all items rather than only those the event gate chose to revise, v0.4 measures the mechanism layer the v0.3 result implicated. The v0.4 numbers --- shadow revision strictly beats draft at $g=+0.65$ and a learned gate beats event-gated at F1${}=0.65$ vs.\ $0.52$ --- are exactly the kind of finding the v0.3 pilot motivated and could not, at its sample size, deliver.

\paragraph{The v0.3 results were left on a branch and never merged before v0.4 began.} An honest reader of the repository will notice that the Phase~7 results that drive every number above were left for several days on the \texttt{v0.4-mechanism-study} branch and never merged to \texttt{main}. We address this directly. The v0.3 SPEC closed with a negative-result obligation: any pilot that failed to find the predicted directional effect must be reported as that finding, not framed as a methodological caveat. The v0.3 prose acknowledged the obligation but did not enact it; the prose framed the inconclusive headline as a sample-size limitation and set up additional control arms for a v0.4 rerun. v0.4 inherits the obligation and discharges it: the headline cascade-versus-bare result is inconclusive at $n=25$, and we say so in the abstract; the load-bearing mechanism findings (H8a, H8c, H9, H8b) are foregrounded for what they actually are; the speculative scope-cuts (chandas-aware scorer, Hopfield \iast{\=alayavij\~n\=ana} substrate, learned gate trained on more data) are deferred to v0.5 with explicit falsification criteria. The unmerged-state evidence-then-prose discipline is, in retrospect, a feature of the pipeline rather than a bug --- the artefacts on the branch are the truth, and v0.4 is the prose rebuild around them.

\paragraph{Cascade $K$ deferred from $K{=}4$ to $K{=}3$.} The pilot used $K{=}3$ to keep total Haiku spend under cap. v0.1 ran at $K{=}4$. The v0.3 run is scoped at $K{=}3$; a follow-up at $K{=}4$ on a properly powered $n{=}20$/domain pull will eliminate this confound.

\paragraph{Scoring functions are local proxies, not gold standards.} We deliberately use local deterministic scorers to keep the contrast fair across arms. They are, however, proxies: POEMetric is a more nuanced rubric than what our \texttt{score\_poetry\_gen} function captures, and the Wittgenstein aspect-multiplicity score is constructed from cosine to author-supplied aspect targets rather than from human raters. A follow-up with human raters (or a frontier-model judge) is necessary before the absolute scores can be interpreted as POEMetric scores in the original sense; our \emph{contrasts} between arms remain valid because the same scorer is applied uniformly to all arms.

\paragraph{Prompt-engineering confound.} The Claude Code CLI surrounds every prompt with a Claude Code system context (visible in the cache-creation token count). This system context could in principle help or hurt the Haiku arm in domain-specific ways. We held the user prompt constant across arms and used Claude Code's own \texttt{-p} non-interactive mode (the most LLM-API-like surface), but we cannot prove the system prompt does not bias the Haiku arm's outputs.

\paragraph{Model substrate confound.} Qwen2-1.5B-Instruct is the only local LLM we tested. The PCE design is substrate-agnostic, but a wider sweep (Qwen2-7B, Llama-3-8B, Mistral-Nemo) is needed to confirm the cascade contributes consistently across substrate scales.

\paragraph{Internal-validity vs external-validity.} The H6 test (within-PCE event vs no-event) is a within-condition test and does not establish that the recursive layer would help under a different scoring function or a different prompt distribution. We report it as an \emph{internal} validity check, not as a generalisable claim.

\paragraph{Single-author authorship of the benchmark items.} The Wittgenstein aspect-shift items are author-constructed. We avoided cherry-picking by stratifying across canonical poetic devices (metaphor, synecdoche, allegory, etc.), but the items have not been reviewed by independent literary scholars. We publish them in Appendix~\ref{app:benchmark_items} so the community can replace or augment them.
