\section{Related Work}\label{sec:related}

\subsection{Computational instantiations of Pratyabhij\~n\=a}

There is a small but growing literature that takes Indic philosophical architectures as more than metaphor for AI systems.~\citet{RatieFlamandLeavitt2025} reconstruct \iast{vimar\'sa} as a recursive consciousness operator. ~\citet{NambhatraDahlquist2025} formalise \iast{kriy\=a-yoga} as embodied agentive practice in a generative-model setting. \citet{CONNECTOM2024,Suriya2025VimarsaConsciousness} explicitly propose \iast{vimar\'sa} as the recursive self-knowledge step for synthetic agents. Our work differs in three respects: (i) we ground the cascade in measurable, reproducible operators on a real LLM substrate rather than a connectionist theoretical sketch; (ii) we make the \iast{vimar\'sa} loop a falsifiable trigger under a free-energy criterion, not a generic recurrent module; (iii) we publish the plugin and benchmark artefacts so any third party can replay the contrast. This reflects the Pratyaksha (Skt.\ \emph{pratyak\d{s}a}, ``perception''/``directly evidenced'') stance --- the witness invariant of the Pratyaksha Context Engineering Harness~\citep{PratyakshaContextEng2025} --- where every claim is anchored to externally verifiable artefacts.

\subsection{Active inference and Bayesian Model Reduction}

\citet{FristonPenny2011} introduced Bayesian model reduction (BMR) as a method for posterior switching under nested models, shown to recover sparse posteriors from rich ones via parameter pruning. \citet{Friston2018BayesianModelReductionArxiv} generalised this to free-energy minimisation in active inference, and \citet{Friston2017ActiveInferenceCuriosity} explicitly link BMR to insight: a sudden re-organisation of the posterior is the formal correlate of the ``aha'' moment. \texttt{pymdp} \citep{Heins2022pymdp} provides a discrete-state-space implementation. \citet{vanDeMaele2024NeurIPS} demonstrate that BMR over a hierarchical generative model produces interpretable abstraction during sleep-like consolidation. We use BMR as the formal substrate of the \iast{j\~n\=ana} operator (posterior over candidates) and the SWS/REM consolidation cycle.

\subsection{LLM creativity benchmarks}

POEMetric~\citep{Ye2025POEMetric} provides a 6-axis poetry-generation rubric (creativity, lexical diversity, idiosyncrasy, emotional resonance, literary devices, imagery) with strong inter-rater agreement. CreativityPrism~\citep{CreativityPrism2025} aggregates AUT, TTCT, and other divergent-thinking items under a Quality$\cdot$Novelty$\cdot$Diversity decomposition. BIG-Bench Hard~\citep{Suzgun2022BBH} contributes scientific-reasoning probes; MacGyverBench~\citep{Tian2024MacGyverBench} contributes constraint-satisfaction creativity items. The Wittgenstein aspect-shift task is a custom $n=20$ stratified slice we construct in \S\ref{sec:benchmarks} on principled grounds, since no public benchmark explicitly tests \emph{simultaneously-coexisting} interpretations of a single line of poetry.

\subsection{Plugin architectures for LLMs}

The Cursor and Claude Code ecosystems have converged on a manifest-driven plugin model with MCP tools, slash commands, hooks, agents, and skills. We follow the same conventions used by \texttt{attractor-flow} \citep{AttractorFlow2025} (which couples a real attractor-network engine to MCP) so that the PCE plugin can be installed and audited without any framework-specific shim. Our plugin shape (15 MCP tools, 5 skills, 5 agents, 5 commands, 3 hooks) is designed for a single-session, real-inference creative-cognition workflow.
