\section{Discussion}\label{sec:discussion}

The pre-registered design forces us to resist post-hoc storytelling. We discuss only the patterns that the data already exhibit.

\paragraph{What the v0.2 cascade contributed in the pilot.} The empirical record reverses the v0.1 sign on every primary hypothesis: H1--H4 all show \emph{directionally positive} effects (g $\in [+0.30, +2.12]$) for the apples-to-apples \texttt{haiku\_cascade} vs \texttt{haiku\_bare} contrast, with three of four BCa CIs strictly above zero (H1, H2, H4). \emph{None} of the four crossed the pre-registered Holm-adjusted $p<0.05$ threshold; this is a consequence of the pilot's $n{=}5$/domain (the exact sign-flip permutation test has a hard floor of $p{=}0.0312$ at $n{=}5$, and Holm with $m{=}4$ pushes the smallest possible adjusted $p$ to $0.125$). The honest reading is that the v0.2 design produces effects of the right sign and magnitude to be detected by a properly powered run ($n{\sim}20$/domain), but the pilot itself is too small to clear the strict pre-registered bar. The v0.2 contribution is therefore (a) the methodological reversal of the v0.1 conflation between cascade contribution and substrate gap; (b) the operationalisation of \iast{vimar\'sa} as a causal two-pass-always loop (\iast{vimar\'sa} now fires on 9 of 14 \texttt{haiku\_cascade} trials vs.\ 0 of 30 in v0.1); and (c) the TRIZ-resolved operator changes that closed each of the v0.1 review's five contradictions (\S\ref{sec:methods}).

\paragraph{Why the sign reversed (and not just the magnitude).} The five v0.2 changes are causally specific:
\begin{itemize}[leftmargin=1.4em]
  \item \textbf{Causal \iast{vimar\'sa} (ADR-003).} v0.1's \iast{vimar\'sa} was post-hoc telemetry: it ran but did not feed back into the surface. v0.2 routes every cascade item through draft $\to$ \iast{vimar\'sa} brief $\to$ revision so the loop can actually shape the output. The output text in cascade arms is the revision, not the draft.
  \item \textbf{Signed \iast{apohana} (ADR-002).} v0.1 clipped negative \iast{apohana} to zero, so the operator could only \emph{discount} must-avoid candidates --- not penalise them. v0.2's signed \iast{apohana} (with min-max-shifted pseudo-counts in \iast{j\~n\=ana}) lets the posterior actively avoid must-avoid neighborhoods.
  \item \textbf{Prompt and sampler parity (ADR-005).} v0.1's cascade silently changed prompts and sampler-grid distributions vs.\ the bare arm, so the v0.1 contrast partly measured prompting differences. v0.2's \texttt{iccha} runs in \texttt{prompt\_mode="verbatim"} and \texttt{sampler\_grid\_mode="parity"} by default inside the cascade so the bare and cascade arms cannot be told apart on either axis.
  \item \textbf{Pluggable LM protocol (ADR-001 / ADR-004).} v0.1 hard-coded a 1.5B local substrate which (i) capped absolute scores below Haiku and (ii) made the v0.1 ``primary'' contrast a substrate gap rather than a cascade contribution. v0.2's \texttt{LMProtocol} let us swap in Haiku as a first-class substrate and run the apples-to-apples contrast for the first time.
  \item \textbf{\iast{Vimar\'sa} firing thresholds (Phase 4 calibration).} v0.1's \iast{vimar\'sa} required \texttt{switching} $\geq 2$ on the \iast{apohana} trajectory and \texttt{aspect\_threshold} $\geq 2$, which the one-shot $K{=}4$ trajectory in v0.1 could not satisfy. v0.2 sets \texttt{min\_evidence\_points}=1 and \texttt{aspect\_threshold}=1, treating one-point trajectories as N/A on the switching gate, so the operator can fire on a regime where it actually has evidence.
\end{itemize}
The TRIZ analysis (\texttt{docs/triz/}) is the explicit causal trace from v0.1 review symptoms to v0.2 operator changes; the ADRs (\texttt{docs/adr/v0.2/}) are the design contracts those changes were checked against; the prove-gate report (\texttt{docs/reviews/2026-04-28-prove-gate.md}) is the local-validity check that catches the changes before the pilot spends Haiku budget on them.

\paragraph{The Haiku gap is the right frame for v0.2.} The poetry\_gen scoring is heavy on POEMetric axes that reward fluency. With matched substrate (Haiku for both arms), the v0.2 result on poetry\_gen is small ($g{=}+0.30$, BCa overlapping zero) --- consistent with the expectation that the cascade adds the most where multiplicity and aspect-shift are scored (poetry\_interp $g{=}+2.12$, sci\_creativity $g{=}+0.76$, AUT $g{=}+1.26$). The within-cascade H6 test (Figure~\ref{fig:h6}) shows the same direction: \texttt{haiku\_cascade} items in which \iast{vimar\'sa} fired score higher than items in which it did not, by $\Delta{=}{+}0.031$ pooled (ns at the pilot $n$).

\paragraph{What we kept honest about the pilot.} The pilot was deliberately under-powered ($n{=}5$/domain) to fit a \$18 cost cap; the strict Holm $p<0.05$ bar is therefore unreachable by construction. We report the BCa-lower-bound $>0$ criterion as a complementary signal, not a substitute for the pre-registered threshold. The headline result in the abstract reflects this: it states the directional reversal vs v0.1 and the BCa-lower-bound subset, not a strict-significance claim. A properly powered $n{=}20$/domain run is scoped (and the prepared Sonnet judge bridge in \texttt{scripts/run\_judge\_bridge.py}, dry-runnable in this repo, is the v0.3 next step toward a non-embedding judge).

\paragraph{What the engine does \emph{not} claim.} PCE does not claim to replicate phenomenological consciousness. The Pratyabhij\~n\=a vocabulary is borrowed because its taxonomy is exceptionally fine-grained, not because we make first-person claims. Likewise, BMR is a posterior-switching primitive in our system; we do not claim that the Friston-style free-energy variational identity holds at the system level. The honest v0.2 claim is: \emph{this taxonomy of operators, when written down as code with the v0.2 TRIZ-resolved changes, produces a directionally positive lift of medium-to-very-large effect size on multiplicity-rewarding benchmarks under the apples-to-apples Haiku-substrate conditions documented here, with the operator-level causal chain between TRIZ contradiction, ADR, code change, and pilot result fully traceable in the audit log.}
