\section{Limitations and Threats to Validity}\label{sec:limitations}

\paragraph{Pilot is under-powered by design.} The v0.2 pilot was scoped to fit a \$18 Haiku budget at $n{=}5$/domain ($N_{\text{paired}}{=}19$), $K{=}3$, max\_tokens$=150$. With $n{=}5$ the exact sign-flip permutation test has a hard floor of $p{=}0.0312$ and Holm--Bonferroni with $m{=}4$ floors the smallest possible adjusted $p$ at $0.125$. \emph{No primary hypothesis (H1--H4) can cross the pre-registered Holm $p<0.05$ bar at this sample size}, regardless of effect direction or magnitude. The pilot's role is therefore to validate the \emph{direction} and \emph{calibration} of the v0.2 design, not to deliver a strict-significance claim. A properly powered run at $n{\sim}20$/domain is the v0.3 next step.

\paragraph{Local-cascade arm deferred to v0.3.} The pilot's local-cascade arm (\texttt{local\_cascade}) was excluded from the run after a wall-clock probe showed a single item taking $\sim$6 minutes on the pilot host (a single Apple-Silicon laptop) at $K{=}3$, max\_tokens$=150$ --- which would have pushed the pilot past 3 hours. The local-ablation contrast is therefore unmeasured in this paper and is the most important methodological gap; the v0.3 plan executes it on a GPU host. The two remaining v0.2 hypotheses (H7 substrate-portability and H8 revision-causality) are also deferred to v0.3 for the same reason.

\paragraph{\iast{Vimar\'sa} firing rate is now non-zero, but the H6 contrast is under-powered.} v0.2's two-pass-always cascade observed \iast{vimar\'sa} firing on $9/14$ \texttt{haiku\_cascade} trials --- a structural fix relative to v0.1's 0/30 firing rate. The H6 within-cascade contrast (fired vs not) shows the right direction ($\Delta{=}{+}0.031$) but at $n_{\text{fired}}{=}9, n_{\text{not}}{=}5$ the Mann--Whitney $U$ has $p{=}0.30$. We treat the firing-rate jump as a calibration finding (the v0.2 \texttt{min\_evidence\_points}=1, \texttt{aspect\_threshold}=1, and N/A switching gate fixes work as intended) and the H6 effect as directional-but-underpowered.

\paragraph{Cascade $K$ deferred from $K{=}4$ to $K{=}3$.} The pilot used $K{=}3$ to keep total Haiku spend under cap. v0.1 ran at $K{=}4$. The v0.3 run is scoped at $K{=}4$ on a properly powered $n{=}20$/domain pull and will eliminate this confound.

\paragraph{Scoring functions are local proxies, not gold standards.} We deliberately use local deterministic scorers to keep the contrast fair across arms. They are, however, proxies: POEMetric is a more nuanced rubric than what our \texttt{score\_poetry\_gen} function captures, and the Wittgenstein aspect-multiplicity score is constructed from cosine to author-supplied aspect targets rather than from human raters. A follow-up with human raters (or a frontier-model judge) is necessary before the absolute scores can be interpreted as POEMetric scores in the original sense; our \emph{contrasts} between arms remain valid because the same scorer is applied uniformly to all arms.

\paragraph{Prompt-engineering confound.} The Claude Code CLI surrounds every prompt with a Claude Code system context (visible in the cache-creation token count). This system context could in principle help or hurt the Haiku arm in domain-specific ways. We held the user prompt constant across arms and used Claude Code's own \texttt{-p} non-interactive mode (the most LLM-API-like surface), but we cannot prove the system prompt does not bias the Haiku arm's outputs.

\paragraph{Model substrate confound.} Qwen2-1.5B-Instruct is the only local LLM we tested. The PCE design is substrate-agnostic, but a wider sweep (Qwen2-7B, Llama-3-8B, Mistral-Nemo) is needed to confirm the cascade contributes consistently across substrate scales.

\paragraph{Internal-validity vs external-validity.} The H6 test (within-PCE event vs no-event) is a within-condition test and does not establish that the recursive layer would help under a different scoring function or a different prompt distribution. We report it as an \emph{internal} validity check, not as a generalisable claim.

\paragraph{Single-author authorship of the benchmark items.} The Wittgenstein aspect-shift items are author-constructed. We avoided cherry-picking by stratifying across canonical poetic devices (metaphor, synecdoche, allegory, etc.), but the items have not been reviewed by independent literary scholars. We publish them in Appendix~\ref{app:benchmark_items} so the community can replace or augment them.
