\section{Limitations and Threats to Validity}\label{sec:limitations}

\paragraph{Substrate gap is the dominant confound.} Comparing a 1.5B-parameter local Qwen2 substrate (with or without the cascade) to Anthropic's Haiku is structurally unfavourable to the substrate, and the absolute composite gap in Figure~\ref{fig:per_domain_box} is consistent with that gap rather than with a defect of the architecture. The cascade-vs-bare sensitivity contrast (Section~\ref{sec:sensitivity}) holds the substrate constant and shows small, non-directional effects: this is the methodologically clean comparison and we recommend it as the primary test for replications.

\paragraph{\iast{Vimar\'sa} did not fire (n\_fired = 0/30).} The recursive layer is the central novel contribution and was \emph{dormant} for the full duration of the run. Two specific knobs are responsible: (a) the predicted-token-on-top criterion compares the LLM's argmax token to a small held-out probe slot, and on Qwen2-1.5B at $\tau{=}0.7, p{=}0.95$ the argmax rarely matched; (b) the Bayesian model-reduction trigger uses a $\Delta\mathcal{F}>2$ nat threshold that is conservative by design. Future runs should re-tune both before the recursive contribution can be evaluated.

\paragraph{Cascade budget under-powered.} We ran with $K{=}4$ candidates and \texttt{max\_tokens}=120 to fit in MPS wall-clock budget. Both choices restrict the regime in which \iast{ic\d{c}h\=a}'s candidate diversity matters. A re-run at $K{=}16,\ \texttt{max\_tokens}{=}512$ would cost $\sim$8$\times$ the wall-clock but is the principled next experiment.

\paragraph{Sample size.} Per-domain $n$ is bounded above by the wall-clock budget of running the cascade on a single MPS device against the Claude Code CLI's quota. The a-priori power table (Figure~\ref{fig:power}) shows the smaller-$n$ domains are under-powered for $g=0.5$, and we flag every conclusion that rests on under-powered evidence as ``provisional''.

\paragraph{Scoring functions are local proxies, not gold standards.} We deliberately use local deterministic scorers to keep the contrast fair across arms. They are, however, proxies: POEMetric is a more nuanced rubric than what our \texttt{score\_poetry\_gen} function captures, and the Wittgenstein aspect-multiplicity score is constructed from cosine to author-supplied aspect targets rather than from human raters. A follow-up with human raters (or a frontier-model judge) is necessary before the absolute scores can be interpreted as POEMetric scores in the original sense; our \emph{contrasts} between arms remain valid because the same scorer is applied uniformly to all arms.

\paragraph{Prompt-engineering confound.} The Claude Code CLI surrounds every prompt with a Claude Code system context (visible in the cache-creation token count). This system context could in principle help or hurt the Haiku arm in domain-specific ways. We held the user prompt constant across arms and used Claude Code's own \texttt{-p} non-interactive mode (the most LLM-API-like surface), but we cannot prove the system prompt does not bias the Haiku arm's outputs.

\paragraph{Model substrate confound.} Qwen2-1.5B-Instruct is the only local LLM we tested. The PCE design is substrate-agnostic, but a wider sweep (Qwen2-7B, Llama-3-8B, Mistral-Nemo) is needed to confirm the cascade contributes consistently across substrate scales.

\paragraph{Internal-validity vs external-validity.} The H6 test (within-PCE event vs no-event) is a within-condition test and does not establish that the recursive layer would help under a different scoring function or a different prompt distribution. We report it as an \emph{internal} validity check, not as a generalisable claim.

\paragraph{Single-author authorship of the benchmark items.} The Wittgenstein aspect-shift items are author-constructed. We avoided cherry-picking by stratifying across canonical poetic devices (metaphor, synecdoche, allegory, etc.), but the items have not been reviewed by independent literary scholars. We publish them in Appendix~\ref{app:benchmark_items} so the community can replace or augment them.
