\section{Plugin Surface}\label{sec:plugin}

PCE is delivered as a Claude Code plugin (manifest at \texttt{plugin/.claude-plugin/plugin.json}). Its four user-visible surfaces are:

\paragraph{Skills (5).} Workflow-level guidance read by an agent before acting: \texttt{pce-poetry-generation}, \texttt{pce-poetry-interpretation}, \texttt{pce-divergent-thinking}, \texttt{pce-scientific-creativity}, \texttt{pce-vimarsa-self-reflection}.

\paragraph{Agents (5).} Specialised personas: \texttt{pce-poet}, \texttt{pce-interpreter}, \texttt{pce-ideator}, \texttt{pce-scientist}, \texttt{pce-vimarsa-auditor}.

\paragraph{Slash commands (5).} \texttt{/pce-compose}, \texttt{/pce-interpret}, \texttt{/pce-aut}, \texttt{/pce-bbh}, \texttt{/pce-trace} (the last surfaces the audit trail of a recent run).

\paragraph{Hooks (3).} \texttt{SessionStart} (initialises the engine and HF auth), \texttt{PreToolUse} (audit-stamps every PCE MCP call), \texttt{PostToolUse} (writes the consolidation tick).

\paragraph{MCP tools (19 in v0.3).} The plugin exposes 19 FastMCP tools wrapping the engine. The 15 v0.1/v0.2 tools (\texttt{pce.cit}, \texttt{pce.ananda}, \texttt{pce.iccha}, \texttt{pce.apohana}, \texttt{pce.jnana}, \texttt{pce.kriya}, \texttt{pce.vimarsa}, \texttt{pce.cascade}, \texttt{pce.embed}, \texttt{pce.lm.generate}, \texttt{pce.lm.entropy}, \texttt{pce.store.add}, \texttt{pce.store.recall}, \texttt{pce.store.consolidate\_sws}, \texttt{pce.store.consolidate\_rem}) are preserved for backward compatibility. v0.3 adds: \texttt{pce.pce\_cascade} (the v0.3 arm-switchable cascade taking \texttt{commit\_policy}, \texttt{cit\_temperature}, \texttt{use\_storehouse}, \texttt{hopfield\_weight}, and the new control-arm names \texttt{haiku\_bare\_2K} and \texttt{haiku\_generic\_revise}), \texttt{pce.haiku\_clean\_substrate\_probe} (live \texttt{IntegrityProbe} against the inner \texttt{claude --print} subprocess; returns \texttt{passed}, \texttt{leak\_matches}, \texttt{positive\_hint}, and the cached \texttt{env\_hash}/\texttt{flags\_hash}), and \texttt{pce.hopfield\_state} (introspect the active-inference \iast{\=alayavij\~n\=ana}: per-domain pattern counts and recent L2 norms).

A smoke-test harness in \texttt{scripts/smoke\_plugin.py} validates the manifest, the component counts (19 tools, 5 skills, 5 agents, 5 slash commands, 3 hooks), and exercises every MCP tool in-process. A separate \texttt{scripts/verify\_outer\_host\_loads\_pce.py} guards against outer-host regressions caused by the v0.3 inner-substrate hardening: only the \emph{inner} Haiku CLI subprocess runs in the clean environment; the outer Claude Code host that loads PCE itself must keep its plugins, skills, and MCP servers intact. The phase-6 audit log is published at \texttt{audit/phase6\_v0.3/}.
