\section{Statistical Protocol}\label{app:stats}

The statistical pipeline lives in \texttt{benchmarks/stats.py} and is exercised by the synthetic-data tests in \texttt{tests/test\_stats.py}. We re-state it here as a self-contained appendix.

\subsection{Paired permutation}
For each hypothesis we form the per-item paired delta $d_i = T_i - C_i$ where $T$ is the treatment composite and $C$ is the control composite. We test $H_0: \mathbb{E}[d] = 0$ vs $H_1: \mathbb{E}[d] > 0$ by sign-flip permutation: under $H_0$, the sign of $d_i$ is exchangeable. We enumerate all $2^n$ sign assignments when $n \leq 18$ (the exact regime), otherwise sample $50\,000$ random sign-vectors. The $p$-value is the proportion of permuted means at least as large as the observed mean, with a $+1$ smoother in the Monte-Carlo regime: $p = (k+1)/(N_{\text{perm}}+1)$.

\subsection{Hedges' $g$}
\[
g = J\,\frac{\bar d}{s_d},\qquad J = 1 - \frac{3}{4n - 5},
\]
where $s_d$ is the unbiased SD of $d$. $J$ is the small-sample bias correction (Borenstein 2009). When $s_d = 0$ we report $g=0$ by convention.

\subsection{BCa bootstrap CI}
$10\,000$ resamples of the paired deltas with replacement; bias-correction $z_0 = \Phi^{-1}\bigl(\#\{b : \bar d^*_b < \bar d\}/B\bigr)$ and acceleration $a$ via jackknife. We use \texttt{scipy.stats.norm.cdf/ppf} for the BCa adjustments and clip the alpha-quantiles into $[0, 1]$ for numerical safety.

\subsection{Wilcoxon signed-rank}
\texttt{scipy.stats.wilcoxon} with \texttt{alternative="greater"}, \texttt{zero\_method="wilcox"}. Reported alongside the permutation $p$ for robustness against extreme-tail behaviour.

\subsection{Holm--Bonferroni}
We adjust $\{p_{H_1}, p_{H_2}, p_{H_3}, p_{H_4}\}$ jointly. Sorted ascending; the $i$-th adjusted value is $\max(\,(m-i)\,p_{(i)},\,p_{(i-1)}^{\text{adj}}\,)$, clamped to $[0, 1]$. H5 and H6 are not double-counted.

\subsection{Power}
Approximate paired-$t$ power under a noncentral-$t$ distribution: $\text{power} = 1 - \Pr(T_{\text{nc}} \leq t_{\text{crit}})$ with $t_{\text{crit}}=t_{\alpha,\,n-1}$ and noncentrality $\lambda = g\sqrt{n}$. We report a-priori (assumed $g=0.5$) and retrospective (observed $g$) power.

\subsection{H6}
Within-PCE event-vs-no-event. Independent-samples Mann--Whitney $U$ (\texttt{alternative="greater"}), Hedges' $g$ for independent samples (pooled SD), and a $10\,000$-resample bootstrap CI for the difference of means. ``Supported'' iff $U$-test $p < 0.05$ \emph{and} the bootstrap CI's lower bound is strictly positive.
