\section{Scoring Functions}\label{app:scoring}

All four scoring functions live in \texttt{benchmarks/scoring.py} and are deterministic. Each returns a \texttt{Score(axes: dict[str, float], composite: float)}. The composite is the unweighted mean of the per-axis scores; we keep the per-axis breakdown in the audit log for the figures in \S\ref{sec:results}.

\subsection{\texttt{score\_poetry\_gen}}
Six axes (creativity, lexical diversity, idiosyncrasy, emotional resonance, literary devices, imagery), all computed from local statistics:
\begin{itemize}
\item \textbf{creativity}: $1 - \cos(\phi(\text{surface}), \phi(\text{topic}))$, normalised to $[0, 1]$.
\item \textbf{lexical diversity}: type/token ratio.
\item \textbf{idiosyncrasy}: 1 minus the maximum cosine to a corpus of 20 ``canonical'' poems on similar topics.
\item \textbf{emotional resonance}: presence of any of $\{$love, fear, hope, grief, joy, longing$\}$ token-stems weighted by sentence count.
\item \textbf{literary devices}: heuristic regex check for alliteration, anaphora, simile-cue (``like'', ``as''), enjambment.
\item \textbf{imagery}: cosine to a small ``imagery vocabulary'' (\{light, shadow, sky, river, leaf, ...\}) cluster centroid.
\end{itemize}

\subsection{\texttt{score\_poetry\_interp}}
Three axes:
\begin{itemize}
\item \textbf{aspect\_count}: count of aspect targets whose cosine to the surface is $\geq 0.40$.
\item \textbf{novelty}: 1 minus the maximum cosine to the retrieval set.
\item \textbf{coverage}: mean cosine across all aspect targets.
\end{itemize}

\subsection{\texttt{score\_aut}}
Three axes:
\begin{itemize}
\item \textbf{creativity}: 1 minus the cosine to the standard-use sentence.
\item \textbf{lexical\_diversity}: type/token ratio.
\item \textbf{feasibility}: presence of feasibility-cue tokens in the candidate set.
\end{itemize}

\subsection{\texttt{score\_sci\_creativity}}
Four axes:
\begin{itemize}
\item \textbf{non\_textbook\_novelty}: 1 minus the cosine to the textbook one-liner.
\item \textbf{framing\_coverage}: mean cosine across the framing targets.
\item \textbf{depth}: log-length normalised to $[0,1]$.
\item \textbf{multi\_framing}: count of framing targets with cosine $\geq 0.35$, divided by the total.
\end{itemize}
